Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-03-18 paperDistance/paper.tex
references

Here we build and use just such a data-driven model.  In particular,
we use all of the \gaia\ \tgas\ data (\citealt{tgas}), matched to the
\tmass\ photometry (\citealt{skrutskie06}), and in the footprint of
the \panstarrs\ Dust Map (CITE GREEN), to make a model of the
noise-free (or low-noise) color--magnitude diagram (\cmd) of stars.
We use an Empirical-Bayes approach known as Extreme Deconvolution (XD;
\citealt{bovy11}).
This method deconvolves heteroskedastic data to derive an
estimate of the noise-free or low-noise distribution that \emph{would
  have} been observed with far better data.
This method has been used in astrophysics
previously to model the velocity distribution of stars in the Solar
Neighborhood (\citealt{hogg05}; \citealt{bovy09}) and to
perform quasar target selection
in \project{Sloan Digital Sky Survey} data (\citealt{xdqso}; \citealt{xdqsoz}).
Its advantage over other deconvolution methods is that it takes as input,
and handles in a principled way, heteroskedastic data.
Its principle disadvantage for the present purposes is that it requires
or implies a strictly Gaussian noise model.
In the below, this requires us to transform the \cmd\ to a space in which
the noise is approximately Gaussian.

The novel use of \xd\ here is to treat its output---the deconvolved \cmd---as
a prior for use in inference of individual-star distances.
These inferences, one per star, provide much more precise parallax, distance,
or absolute magnitude estimates than we get from each star's primary
\tgas\ parallax measurement alone.
That is, we are using the \xd\ model to de-noise the \gaia\ parallax
measurements.
Technically, since \xd\ produces a maximum-marginalized-likelihood estimator
of the deconvolved distribution,
its use as a prior is only approximately valid; it constitutes using the
data twice.
However, since the data set is large, the \xd\ approximation is not bad; it is
sometimes known as the ``empirical Bayes'' method, and is well studied (DWH REFS).

@@

On a different tack, there has been some discussion in the literature
about how to use a parallax responsibly to infer a distance (CITE TRI+CBJ).
These papers use the \gaia\ measurement to construct a parallax likelihood,
and combine this with a distance prior.
While this is sensible and correct, all present methods proposed along
these lines build in informative (and known-to-be-wrong) assumptions
about the line-of-sight distance distribution to \gaia\ stars.
That is, they build in strong and informative assumptions about the Milky
Way.
The methodology presented in this paper makes exceedingly weak assumptions
about the Milky Way, and similarly weak assumptions about the properties of
stars.
The strongest new assumption made in this work is that for every star
in \tgas, there are other, photometrically (and bolometrically)
similar stars.
That is a much weaker assumption than has been made in other Bayesian
distance estimations.

@@

We make the following assumptions. The method we present is correct
and justifiable under these assumptions, each of which is questionable
in its own right. We return to criticize these assumptions---and the
method that flows from them---in the Discussion below.
\begin{itemize}
\item We make an assumption of \emph{stationarity}; that is, that the
  stars measured at lower signal-to-noise have analogs measured at
  higher signal-to-noise.  In detail, this assumption is wrong, of
  course (for example, distant halo stars are different than local
  disk stars), but the stationarity assumption is fairly weak; as long
  as there is support in the high signal-to-noise stars for the types
  of stars seen at low signal-to-noise, the results will not be
  strongly biased, but this assumption is deep and fundamental to this
  work.
\item We assume that we have large numbers of stars, large enough that
  an empirical-Bayes or maximum-marginalized likelihood is a safe
  approximation to full Bayesian inference. This assumption will be
  least true in the least-populated parts of the \cmd. In the extreme
  case of a stellar type with only one (statistically distinct)
  exemplar in the data set (e.g. a white dwarf), this assumption will be maximally wrong.
\item We assume that the \tgas\ parallax uncertainties dominate the
  noise in any estimates of absolute magnitude. We further assume that
  the parallax and color uncertainties are Gaussian in form with
  correctly known variances. In detail, we assume that the [INSERT
    CORRECT TERMINOLOGY FROM GAIA HERE] parallax uncertainty estimates
  are correct. Here we are making use of the fact that, for
  de-noising, under-estimation of measurement noise is conservative
  (it leads to under-deconvolution).
\item We treat the \panstarrs-based three-dimensional dust maps (CITE
  SCHLAFLY) as correct in their median dust estimates, and their
  effects on color and magnitude. As we optimize the empirical-Bayes model, we iteratively update each star's
  dust correction as its updated, inferred distance. Our assumption is that these corrections are
  good enough, and uncertainties in these are not dominant, for either
  color or absolute magnitude.
\item We assume that the \cmd\ can be represented with a mixture of
  Gaussians in a particular transformed space. This mixture we fix
  (with only heuristic analysis) at 128, 512, and 2048 Gaussians in
  different experiments.
\item We make no use of stellar models; our only assumptions about
  stellar physics are generic and implicit (for example, that the
  \cmd\ is somehow smooth and compact). Dust model dependent on stellar models?
\end{itemize}

We would like to infer the true parallax $\varpi_{\true}$ for
each star given its parallax measurement $\varpi$ and associated
uncertainty $\sigma_{\varpi}$ from \tgas.
We assume (as \gaia\ suggests) that the likelihood is Gaussian
\begin{equation}
p(\varpi \given \varpi_{\true}, I) = \mathcal{N}(\varpi \given \varpi_{\true}, \sigma^2_{\varpi}) \quad ,
\label{eq:gaialike}
\end{equation}
where $I$ is a placeholder containing our assumptions, including all
uncertainty estimates, and $\mathcal{N}(x \given \mu, \sigma^2)$ is a
Gaussian function for $x$ with mean $\mu$ and variance $\sigma^2$.
This is the probability of the data given the true parallax, but we
want to flip it around and infer the true parallax, or distance, to
the star given the measured parallax and it's uncertainty. We do this
using probability theory
\begin{equation}
p(\varpi_{\true} \given \varpi, \sigma^2_{\varpi}) = \frac{1}{Z}\,p(\varpi \given \varpi_{\true}, \sigma^2_{\varpi}) \, p(\varpi_{\true}) \quad ,
\label{eq:bayes}
\end{equation}
where $p(\varpi_{\true})$ is a prior probability density (pdf) for the
parallax to the star, and $Z$ is a normalization constant.
The strength of our inference method here lies in how we generate our prior
pdf for the parallax $p(\varpi_{\true})$.
Instead of assuming a functional form for the spatial distribution of
stars in the Milky Way (CITE TRI--CORYN), or relying on stellar
models, we build a noise-deconvolved estimate of the color--magnitude
diagram (\cmd) from the observed colors and absolute magnitudes of our
full set of stars.
This \cmd\ model becomes, for each star---given its observed color and
magnitude (and uncertainties on those)---a prior pdf for the distance
to that star.

>>>>

Here we build and use just such a data-driven model. In particular,
we use all of the \gaia\ \tgas\ data (\citealt{tgas};
these data were generated by the methodology of \citealt{michalik15}), matched to the
\tmass\ photometry (\citealt{skrutskie06}), and in the footprint of
the \panstarrs\ Dust Map (\citealt{green16}), to make a model of the
noise-free (or low-noise) color--magnitude diagram (\cmd) of stars.
We use an Empirical-Bayes approach known as Extreme Deconvolution (XD;
\citealt{bovy11}).
This method deconvolves heteroskedastic data to derive an
estimate of the noise-free or low-noise distribution that \emph{would
  have} been observed with far better data.
This method has been used in astrophysics
previously to model the velocity distribution of stars in the Solar
Neighborhood (\citealt{hogg05}; \citealt{bovy09}) and to
perform quasar target selection
in \project{Sloan Digital Sky Survey} data (\citealt{xdqso}; \citealt{xdqsoz}).
Its advantage over other deconvolution methods is that it takes as input,
and handles in a principled way, heteroskedastic data.
Its principle disadvantage for the present purposes is that it requires
or implies a strictly Gaussian noise model.
In the below, this requires us to transform the \cmd\ to a space in which
the noise is approximately Gaussian.

The novel use of \xd\ here is to treat its output---the deconvolved \cmd---as
a prior for use in inference of individual-star distances.
These inferences, one per star, provide much more precise parallax, distance,
or absolute magnitude estimates than we get from each star's primary
\tgas\ parallax measurement alone.
That is, we are using the \xd\ model to de-noise the \gaia\ parallax
measurements.
Technically, since \xd\ produces a maximum-marginalized-likelihood estimator
of the deconvolved distribution,
its use as a prior is only approximately valid; it constitutes using the
data twice.
However, since the data set is large, the \xd\ approximation is not bad; it is
sometimes known as the ``empirical Bayes'' method, and is well studied\footnote{For a start, try \url{https://en.wikipedia.org/wiki/Empirical_Bayes_method}.}.

@@

On a different tack, there has been some discussion in the literature
about how to use a parallax responsibly to infer a distance (\citealt{astraatmadja16a}; \citealt{astraatmadja16b}).
These papers use the \gaia\ measurement to construct a parallax likelihood,
and combine this with a distance prior.
While this is sensible and correct, all present methods proposed along
these lines build in informative (and known-to-be-wrong) assumptions
about the line-of-sight distance distribution to \gaia\ stars.
That is, they build in strong and informative assumptions about the Milky
Way.
The methodology presented in this paper makes exceedingly weak assumptions
about the Milky Way, and similarly weak assumptions about the properties of
stars.
The strongest new assumption made in this work is that for every star
in \tgas, there are other, photometrically (and bolometrically)
similar stars.
That is a much weaker assumption than has been made in other Bayesian
distance estimations.

@@

We make the following assumptions. The method we present is correct
and justifiable under these assumptions, each of which is questionable
in its own right. We return to criticize these assumptions---and the
method that flows from them---in the Discussion below.
\begin{itemize}
\item We make an assumption of \emph{stationarity}; that is, that the
  stars measured at lower signal-to-noise have analogs measured at
  higher signal-to-noise.  In detail, this assumption is wrong, of
  course (for example, distant halo stars are different than local
  disk stars), but the stationarity assumption is fairly weak; as long
  as there is support in the high signal-to-noise stars for the types
  of stars seen at low signal-to-noise, the results will not be
  strongly biased, but this assumption is deep and fundamental to this
  work.
\item We assume that we have large numbers of stars, large enough that
  an empirical-Bayes or maximum-marginalized likelihood is a safe
  approximation to full Bayesian inference. This assumption will be
  least true in the least-populated parts of the \cmd. In the extreme
  case of a stellar type with only one (statistically distinct)
  exemplar in the data set (e.g. a white dwarf), this assumption will be maximally wrong.
\item We assume that the \tgas\ parallax uncertainties dominate the
  noise in any estimates of absolute magnitude. We further assume that
  the parallax and color uncertainties are Gaussian in form with
  correctly known variances. In detail, we assume that the [INSERT
    CORRECT TERMINOLOGY FROM GAIA HERE] parallax uncertainty estimates
  are correct. Here we are making use of the fact that, for
  de-noising, under-estimation of measurement noise is conservative
  (it leads to under-deconvolution).
\item We treat the \panstarrs-based three-dimensional dust maps (\citealt{green15})
  as correct in their median dust estimates, and their
  effects on color and magnitude. As we optimize the empirical-Bayes model, we iteratively update each star's
  dust correction as its updated, inferred distance. Our assumption is that these corrections are
  good enough, and uncertainties in these are not dominant, for either
  color or absolute magnitude.
\item We assume that the \cmd\ can be represented with a mixture of
  Gaussians in a particular transformed space. This mixture we fix
  (with only heuristic analysis) at 128, 512, and 2048 Gaussians in
  different experiments.
\item We make no use of stellar models; our only assumptions about
  stellar physics are generic and implicit (for example, that the
  \cmd\ is somehow smooth and compact). Dust model dependent on stellar models?
\end{itemize}

We would like to infer the true parallax $\varpi_{\true}$ for
each star given its parallax measurement $\varpi$ and associated
uncertainty $\sigma_{\varpi}$ from \tgas.
We assume (as \gaia\ suggests) that the likelihood is Gaussian
\begin{equation}
p(\varpi \given \varpi_{\true}, I) = \mathcal{N}(\varpi \given \varpi_{\true}, \sigma^2_{\varpi}) \quad ,
\label{eq:gaialike}
\end{equation}
where $I$ is a placeholder containing our assumptions, including all
uncertainty estimates, and $\mathcal{N}(x \given \mu, \sigma^2)$ is a
Gaussian function for $x$ with mean $\mu$ and variance $\sigma^2$.
This is the probability of the data given the true parallax, but we
want to flip it around and infer the true parallax, or distance, to
the star given the measured parallax and it's uncertainty. We do this
using probability theory
\begin{equation}
p(\varpi_{\true} \given \varpi, \sigma^2_{\varpi}) = \frac{1}{Z}\,p(\varpi \given \varpi_{\true}, \sigma^2_{\varpi}) \, p(\varpi_{\true}) \quad ,
\label{eq:bayes}
\end{equation}
where $p(\varpi_{\true})$ is a prior probability density (pdf) for the
parallax to the star, and $Z$ is a normalization constant.
The strength of our inference method here lies in how we generate our prior
pdf for the parallax $p(\varpi_{\true})$.
Instead of assuming a functional form for the spatial distribution of
stars in the Milky Way (\citealt{astraatmadja16b}), or relying on stellar
models (\citealt{gaia16}), we build a noise-deconvolved estimate of the color--magnitude
diagram (\cmd) from the observed colors and absolute magnitudes of our
full set of stars.
This \cmd\ model becomes, for each star---given its observed color and
magnitude (and uncertainties on those)---a prior pdf for the distance
to that star.
